---
title: "Empirical Power For The Sign-Test,T-test and The Wilcoxon Signed Rank Test"
author: "Acquah Theophilus B.K."
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    toc: yes
    fig_caption: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```


```{r, include=FALSE}
install.packages('tinytex')
library(tinytex)
#tinytex::tlmgr_install("fancyhdr")
#tinytex::reinstall_tinytex()
```

\vspace{10pt}
```{r, include=FALSE}
# Load required packages
if (!require(pwr)) install.packages("pwr")
library(pwr)
if (!require(coin)) install.packages("coin")
library(coin)

if (!require(VGAM)) install.packages("VGAM")
library(VGAM)

options(install.packages.check.source = "no")
install.packages("dplyr", type = "binary");library(dplyr)

#install.packages("dplyr", dependencies = TRUE);library(dplyr)
#install.packages("dplyr");library(dplyr)
install.packages("tidyr");library(tidyr)
install.packages("knitr");library(knitr)

```

```{r, include=FALSE}
# Load necessary libraries
if (!require(VGAM)) install.packages("VGAM")
if (!require(dplyr)) install.packages("dplyr")
library(VGAM)
library(dplyr)

# Load necessary libraries
if (!require(VGAM)) install.packages("VGAM")
if (!require(dplyr)) install.packages("dplyr")
library(VGAM)
library(dplyr)
```


Let $X_1, \ldots, X_n$ be a random sample from an unknown continuous distribution
which is symmetric about $\theta$ with distribution function $G_0(x) = F(x-\theta)$.
That is $P_0(X \leq x) = F(x-\theta)$. Now, we test
\[
H_0 : \theta = 0 \quad \text{versus} \quad H_1 : \theta > 0. \tag{1}
\]
Note that this model is indexed by two unknown parameters, $\theta$ and the 
underlying distribution $F(.)$. Here $F$ is an unknown function which is symmetric
about zero. That is $F(x) = F(-x), x\in \mathbb{R}$.

\begin{enumerate}
    \item \textbf{sign-test}\\
    The sign test, with test statistic $B = [ \text{number of } X_i's > 0]$, is
    appropriate for this problem. Under $H_0$, $B$ is a binomial random variable 
    with probability of success equal to 0.5. For any $\theta>0$, $B$ is a 
    binomial random variable with probability of success equal to
    \[
    1 - G_0(0) = 1 - F(-\theta) = F(\theta). \tag{2}
    \]
    The last equality follows from the fact that $F$ is a symmetric distribution.
    So if the rule rejects whenever $B\geq b_0$, the power function is calculated by
    \[
    p((\theta, F) = \sum_{b=b_0}^{n} \binom{n}{b} F(\theta)^b \left[1 - F(\theta)\right]^{n-b}.
    \]

    \item \textbf{t-test}\\
    The one-sample Student t-test for this location problem. We reject $H_0$ if
    \[
    T^+ = \frac{\bar{X}}{S/\sqrt{n}}
    \]
    is greater than $t_{1-\alpha, n-1}$, the upper $100(1 - \alpha)$th quantile 
    of a t-distribution with $n - 1$ degree of freedom. If the underlying 
    distribution is normal with mean $\theta$ and variance $\sigma^2$, the power
    function of the test based on $T^+$ is
    \[
    p(T^+ > t(\alpha, n - 1)|\theta, \sigma) = P\left(\frac{\bar{X}}{S/\sqrt{n}} > 
    t(\alpha, n - 1)|\theta, \sigma\right) \tag{3}
    \]
    \[
    = P\left(\frac{\frac{\bar{X}-\theta}{\sigma/\sqrt{n}}+\frac{\theta}{\sigma/\sqrt{n}}}{(s/\sqrt{n})} >         t(\alpha, n - 1)|\theta, \sigma\right) \tag{4}
    \]
    \[
    = P\left(T^* > t(\alpha, n - 1)|n - 1, \frac{\theta}{\sigma/\sqrt{n}}\right), \tag{5}
    \]
    where $T^*$ has a non-central t-distribution with degree of freedom $n - 1$,non-centrality 
    parameter $\frac{\theta}{\sigma/\sqrt{n}} =\sqrt{n}\theta/\sigma$
    
\item \textbf{Wilcoxon Signed Rank Test}\\
The power function of the Wilcoxon signed rank test is more complex, and we use 
simulation to calculate the empirical power function. You can calculate empirical 
power for the sign-test and t-test, if you do not want to use the exact power 
function given above.Replace the distribution function $F$ with the following 
symmetric distributions:

\begin{enumerate}
    \item Uniform distribution with mean 0 and variance 1,
    \item Normal distribution with mean 0 and variance 1, and
    \item Double exponential distribution with mean 0 and variance 1.
\end{enumerate}

Compute the empirical power for the sign-test, t-test, and the Wilcoxon signed 
rank test based on generating random samples of sizes 10, 15, and 20 from the 
above distributions.

Please refer to the attached PDF files on Canvas for more details on power function
and empirical power function.

\end{enumerate}

\textcolor{red}{CODE IMPLEMENTATION}

>    The following cumulative distribution functions (CDFs) are defined for 
>    distributions that are symmetric around zero and standardized to have a variance of 1:
\begin{enumerate}
 >   \item \textbf{Uniform distribution:} The uniform distribution is defined over an interval 
 >   $[a, b]$. To achieve a mean of 0 and a variance of 1, we set the bounds to $a = -\sqrt{3}$ 
 >   and $b = \sqrt{3}$ since the variance of a uniform distribution over $[a, b]$ is 
 >   $\frac{(b-a)^2}{12}$.
>    \item \textbf{Normal distribution:} The normal distribution is parameterized by its mean 
>    ($\mu$) and standard deviation ($\sigma$). A standard normal distribution, which has a mean 
>    of 0 and a standard deviation of 1, naturally has a variance of 1.
>   \item \textbf{Double exponential (Laplace) distribution:} The variance of the Laplace distribution
>   is given by $2 \cdot \text{scale}^2$. To obtain a variance of 1, the scale parameter is set to 
>   $\sqrt{0.5}$.
\end{enumerate}


\textcolor{red}{Define the cumulative distribution functions 'F' for the specified distributions}

```{r}
# Define the cumulative distribution functions 'F' for the specified distributions

F_uniform <- function(x) punif(x, min = -sqrt(3), max = sqrt(3))
F_normal <- function(x) pnorm(x, mean = 0, sd = 1)
F_double_exponential <- function(x) plaplace(x, location = 0, scale = sqrt(0.5))
```

> Define the power function for the sign test using a symmetric distribution

>The \texttt{switch} function in R evaluates an expression and matches the 
>result against a list of possible cases to execute corresponding code blocks. 
For example:
\begin{verbatim}
switch(expression,
       case1 = code_for_case1,
       case2 = code_for_case2,
       ...
       caseN = code_for_caseN)
\end{verbatim}
Here, \texttt{expression} is compared with \texttt{case1}, 
\texttt{case2}, ..., \texttt{caseN}. If there is a match, the code associated 
with that case is executed.

```{r}
# Define the power function for the sign test using a symmetric distribution
sign_test_power <- function(n, theta, F, alpha = 0.05){
  b0 <- qbinom(alpha, size = n, prob = 0.5, lower.tail = FALSE)
  sum(sapply(b0:n, function(b) choose(n, b) * F(theta)^b * (1 - F(theta))^(n - b)))
}

# Function to calculate empirical power using simulation
compute_empirical_power <- function(n, theta_over_sigma, dist_name,
                                    nsim = 5000, alpha = 0.05) {
  power_sign_test <- numeric(nsim)
  F <- switch(dist_name,
              uniform = F_uniform,
              normal = F_normal,
              double_exponential = F_double_exponential)
  
for (i in 1:nsim) {
  data <- switch(dist_name,
            uniform = runif(n, min = -sqrt(3), max = sqrt(3)),
            normal = rnorm(n, mean = 0, sd = 1),
            double_exponential = rlaplace(n, location = 0, scale = sqrt(0.5)))
    data <- data + theta_over_sigma
    power_sign_test[i] <- sign_test_power(n, theta_over_sigma, F, alpha)
  }
  mean(power_sign_test)
}
```

\textcolor{red}{Function to calculate empirical power for t-test and Wilcoxon test}
```{r}
# Function to calculate empirical power for t-test and Wilcoxon test
calculate_test_power <- function(n, theta, dist_func, nsim, alpha, test_type) {
  replicate(nsim, {
    data <- dist_func(n) + theta
    if (test_type == "t") {
      p_value <- t.test(data, mu = 0, alternative = 'greater')$p.value
    } else if (test_type == "wilcox") {
      p_value <- wilcox.test(data, mu = 0, alternative = 'greater')$p.value
    }
    p_value <= alpha
  }) %>% mean() * 1000
}
```

\textcolor{red}{Set Parameters, Calculate Power,Run Simulations}
```{r}
# Set parameters
nsim <- 5000
n_values <- c(10, 15, 20)
theta_values <- seq(0, 0.8, by = 0.2)
alpha <- 0.05
set.seed(123)

# Calculate power
results <- expand.grid(n = n_values, theta_over_sigma = theta_values, 
                       dist = c("uniform", "normal", "double_exponential"))
results$T_plus <- NA
results$W_plus <- NA
results$B <- NA

for (i in 1:nrow(results)) {
  n <- results$n[i]
  theta <- results$theta_over_sigma[i]
  dist_name <- results$dist[i]
  
  # Calculate power for t-test
  results$T_plus[i] <- calculate_test_power(n, theta, 
                  switch(dist_name,uniform = function(n) runif(n, min = -sqrt(3),
                  max = sqrt(3)),normal = rnorm,
                  double_exponential = function(n) rlaplace(n, location = 0, 
                  scale = sqrt(0.5))), nsim, alpha, "t")
  
  # Calculate power for Wilcoxon test
  results$W_plus[i] <- calculate_test_power(n, theta, 
                  switch(dist_name,uniform = function(n) runif(n, min = -sqrt(3),
                  max = sqrt(3)),normal = rnorm,
                  double_exponential = function(n) rlaplace(n, location = 0,
                  scale = sqrt(0.5))), nsim, alpha, "wilcox")
  
  # Calculate power for sign test
  results$B[i] <- 1000 * compute_empirical_power(n, theta, dist_name, nsim, alpha)
}

```

\textcolor{red}{Create Table For Uniform Distribution}
```{r}
# Reshape the data for Uniform distribution
uniform_table <- results %>%
  filter(dist == "uniform") %>%
  select(-dist) %>%
  gather(key = "Test", value = "Power",B, T_plus, W_plus) %>%
   mutate(Test = factor(Test, levels = c("T_plus", "W_plus", "B"))) %>%
  spread(key = theta_over_sigma, value = Power) %>%
  mutate(across(where(is.numeric), round, digits = 0))
  

# Print the table for Uniform distribution
kable(uniform_table, caption = "Empirical Power Times 1000 for Uniform Distribution")
```

\textcolor{red}{Create Table For Normal Distribution}
```{r}
# Repeat similar steps for Normal and Double Exponential distributions
normal_table <- results %>%
  filter(dist == "normal") %>%
  select(-dist) %>%
  gather(key = "Test", value = "Power", T_plus, W_plus, B) %>%
  mutate(Test = factor(Test, levels = c("T_plus", "W_plus", "B"))) %>%
  spread(key = theta_over_sigma, value = Power) %>%
  mutate(across(where(is.numeric), round, digits = 0))
  

kable(normal_table, caption = "Empirical Power Times 1000 for Normal Distribution")
```

\textcolor{red}{Create Table For Double Exponential Distribution}
```{r}
double_exp_table <- results %>%
  filter(dist == "double_exponential") %>%
  select(-dist) %>%
  gather(key = "Test", value = "Power", T_plus, W_plus, B) %>%
  mutate(Test = factor(Test, levels = c("T_plus", "W_plus", "B"))) %>%
  spread(key = theta_over_sigma, value = Power) %>%
  mutate(across(where(is.numeric), round, digits = 0))

kable(double_exp_table, caption = "Empirical Power Times 1000 for Double Exponential Distribution")

```







